Revisiting the von Bertalanffy Seasonal Cessational Growth Function of Pauly et al. (1992)

### 2016-05-15 manuscript_v0_2.Rmd
Added references, moved real example up, some editing

---
title: "Revisiting the von Bertalanffy Seasonal Cessational Growth Function of Pauly et al. (1992)"
author: "Derek H. Ogle"
date: "April 29, 2016"
output:
  pdf_document:
    fig_height: 3.5
    fig_width: 4.5
  html_document:
    fig_height: 3.5
    fig_width: 4.5
csl: american-fisheries-society.csl
bibliography: SeasonalGrowth.bib
---

@@

# setup figure, table, and equation captioning
library(captioner)
figcaps <- captioner(prefix="Figure")
tblcaps <- captioner(prefix="Table")
eqncaps <- captioner(prefix="Equation")
eqncaps("vbMain")
eqncaps("vbMain2")
eqncaps("SomersMod")
eqncaps("PaulyMod")
eqncaps("PaulyMod2")

@@

@@

The mean length-at-age for many animals (e.g., CITATIONS) is often modeled with the von Bertalanffy growth function (VBGF; @vonBertalanffy1938).  The parameterization of the VBGF attributable to @BevertonHolt1957 is most common and may be expressed as

@@

Many animals exhibit seasonal oscillations in growth as a response to seasonal changes in environmental factors such as temperature, light, and food supply (CITATIONS).  `r eqncaps("vbMain2",display="cite")` of the traditional VBGF has been modified, usually with a sine function, to model these seasonal oscillations in growth.  The most popular of these modifications, from @HoenigChoudarayHanumara1982 and @Somers1988, is

@@

where $C$ modulates the amplitude of the growth oscillations and corresponds to the proportional decrease in growth at the depth of the oscillation (i.e., "winter"), and $t_{s}$ is the time between time 0 and the start of the convex portion of the first sinusoidal growth oscillation (i.e., the inflection point).  If $C$=0, then there is no seasonal oscillation and `r eqncaps("SomersMod",display="cite")` reduces to  `r eqncaps("vbMain2",display="cite")` and the typical VBGF (`r figcaps("SomersExPlot",display="cite")`).  If $C$=1, then growth completely stops once a year at the "winter-point" ($WP$), whereas values of 0&lt;$C$&lt;1 result in reduced, but not stopped, growth during the winter (`r figcaps("SomersExPlot",display="cite")`).  Note that $WP=t_{s}+\frac{1}{2}$ because the sine function in `r eqncaps("SomersMod",display="cite")` has a period (i.e., the growth period) of one year.  Some confusion has surrounded the use of `r eqncaps("SomersMod",display="cite")`, though @GarciaBerthouetal2012 carefully clarified its form.

@@

Values of $C$&gt;1 (or &lt;0) in  `r eqncaps("SomersMod",display="cite")` allow seasonal decreases in mean length-at-age (`r figcaps("SomersExPlot",display="cite")`).  A decrease in mean length is unlikely for organisms whose skeletons largely preclude shrinkage [@Paulyetal1992], although a seasonal decrease in mean length-at-age is possible if size-dependent overwinter mortality occurs [@GarciaBerthouetal2012].  @Paulyetal1992 modified `r eqncaps("SomersMod",display="cite")` to include a true seasonal no-growth period where mean length was not allowed to decrease and that included a smooth transition of the modeled mean length-at-age into and out of the no-growth period.  Specifically, their modification is

@@

@Paulyetal1992 devised  `r eqncaps("PaulyMod",display="cite")` by assuming $C$=1 and replacing $2\pi$ with $\frac{2\pi}{1-NGT}$ (i.e., restricting the seasonal oscillation to the growth period and noting that $K^{'}$ only operates during the growth period).  Their modification may be described geometrically (though not algorithmically) in two steps.  First, the seasonal growth function with $C$=1 in `r eqncaps("SomersMod",display="cite")` is fit to the observed lengths and ages that have had the cumulative $NGT$ subtracted (i.e., using $t^{'}$).  The growth trajectory is then separated at each $WP$ and horizontal segments that are $NGT$ units long are inserted at these points.  This forms a growth trajectory that smoothly transitions into and out of the no-growth periods (`r figcaps("PaulyExPlot",display="cite")`).

@@

@Paulyetal1992 provided a "diskette" that contained a computer program to estimate the parameters of `r eqncaps("PaulyMod",display="cite")`.  The diskette is difficult (at best) to obtain and the source code is no longer available (D. Pauly, pers. comm.).  @Paulyetal1992 did describe the operations performed by their program, but there is no description of how $t^{'}$ was operationalized.  This is an important step in using `r eqncaps("PaulyMod",display="cite")` because $t^{'}$ is a function of $t$, but it is also a function of $NGT$ and $t_{s}$, which are parameters to be estimated during the model-fitting process.  Thus, the values for $t^{'}$ change with each iteration of the non-linear model-fitting process.

@@

As noted by @Paulyetal1992 the calculation of $t^{'}$ depends on the observed age ($t$) and the cumulative no-growth time prior to $t$.  In practice, the calculation of $t^{'}$ also depends on the position of the no-growth period within a year.  Here, the position of the no-growth period will be defined relative to $WP$ and $NGT$.  The following algorithm may be used to convert from observed ages ($t$) to ages adjusted for cumulative $NGT$ prior to age $t$ ($t^{'}$).

1. Shift the age ($t$) by subtracting the start of the no-growth ($SNG$) period (i.e., $SNG=WP-\frac{NGT}{2}$) from $t$, such that a whole number will represent the start of a no-growth period.  For example, if $SNG$=0.4, then $t$=2.4 will become 2.0 and $t$=2.9 will become 2.5.
1. Subtract the whole number age from the shifted age from Step 1 such that the remaining decimal represents the fraction of a shifted year.  For example, a 0 will result if the shifted age is 2.0 and a 0.5 will result if the shifted age is 2.5.
1. Substract the $NGT$ from the value from the previous step.
1. If the value from the previous step is negative, then the age is within the no-growth period and the negative value should be replaced with a zero.  Otherwise, the positive value represents the amount of time into a growth period.
1. Add the value from the previous step to the total growth time completed (i.e., the product of the number of growth periods completed and the length of the growth period ($1-NGT$)).
1. Compute $t{'}$ by adding back the $SNG$ that was subtracted in Step 1.

@@

# Fitting the Function
## Simulated Data
Simple simulated data are used initially to demonstrate the fitting of `r eqncaps("PaulyMod",display="cite")`.  The data were simulated by randomly selecting 200 real numbers from a uniform distribution between 0 and 5 to serve as observed ages, plugging these ages into `r eqncaps("PaulyMod",display="cite")` to compute a mean length for each age, and then adding a random deviate from a normal distribution with a mean of 0 and a standard deviation of $\sigma$ to each mean length to simulate observed individual lengths.  Eight data sets were simulated for all combinations of "early" ($t_{s}=0.2$) and "late" ($t_{s}=0.8$) maximum growth points, "short" ($NGT=0.2$) and "long" ($NGT=0.5$) no-growth periods, and "low" ($\sigma=0.3$) and "high" ($\sigma=0.5$) individual variabilities.

The `nls()` function from R was then used to estimate parameter values for the nonlinear `r eqncaps("PaulyMod",display="cite")` fit to each simulated data set.  The "port" algorithm was used so that $L_{\infty}$ and $K^{'}$ could be constrained to be positive and $t_{s}$ and $NGT$ could be constrained to be between 0 and 1.  The estimated parameters for each data were consistently very close to the values used to create the data set, with the possible exception of $t_{0}$.  One example, for a "long" $NGT$, "late" $t_{s}$, and low individual variability, is shown in `r figcaps("PaulySimRes",display="cite")`.  The results from fitting `r eqncaps("PaulyMod",display="cite")` to these data sets are not surprising as one would expect the parameter estimates to be very close to the values used to create them, given that `r eqncaps("PaulyMod2",display="cite")` was used to both create and fit the data.

@@

tblcaps("PaulySimRes",paste0("Parameter estimates from fitting ",eqncaps('PaulyMod2',display='cite')," to simulated data sets using $L_{\\infty}$=",Linf,", $K'$=",Kpr,", $t_{0}$=",t0,", and varying values of $t_{s}$, $NGT$, and $\\sigma$."))
```

@@

## Real Data

@Stewartetal2013 examined the growth of 215 Bonito (*Sarda australis*) sampled from commercial landings.  Fork lengths (mm) were measured for each fish and decimal ages were recorded as the number of opaque zones observed on otolith thin sections plus the proportion of the year after the designated birthdate (see @Stewartetal2013 for more detailed methods).  @Stewartetal2013 fit `r eqncaps("SomersMod",display="cite")` to these data but constrained $C$ to not exceed 1.  Their model fit resulted in the boundary condition of $C=1$, which suggested that Bonito ceased to grow at *at least* one point.  This result suggests that `r eqncaps("PaulyMod2",display="cite")` should be fit to these data to determine the length of the no-growth period [@Paulyetal1992].

`r eqncaps("PaulyMod",display="cite")` fit the Bonito data slightly better than `r eqncaps("SomersMod",display="cite")` with slightly lower residual sums-of-squares (RSS) and Akaike Information Criterion (AIC) values.  The length of the no-growth period was estimated to be `r formatC(cfBon2["NGT","Est"],format="f",digits=3)` or `r formatC(cfBon2["NGT","Est"]*100,format="f",digits=1)`% of the year.  The $t_{s}$ parameters were equal, the $L_\infty$ parameters were similar, but the $t_{0}$ parameters differed somewhat between the two models.  Finally, $K$ from `r eqncaps("SomersMod",display="cite")` was equal to $K'$ from `r eqncaps("PaulyMod2",display="cite")` multiplied by $1-NGT$.

```{r echo=FALSE, results="hide"}
## A plot
par(xaxs="i",yaxs="i",mar=c(3,3,0.6,0.6),mgp=c(1.7,.5,0),tcl=-0.3,las=1)
plot(FL~Age,data=bon,pch=19,col="black",
     xlab="Age (years)",ylab="Fork Length (mm)",xlim=c(0,4),ylim=c(0,70))
curve(VBSCGF(x,coef(fitBon2)),from=0,to=4,lwd=4,add=TRUE)
curve(vbSO(x,coef(fitBon1)),from=0,to=4,lwd=2,add=TRUE,col="gray50")

figcaps("BonitoRes",paste0("Fork length at age for Australian Bonito with the best-fit of ",eqncaps('PaulyMod2',display='cite')," (black line) and ",eqncaps('SomersMod',display='cite')," (gray line) superimposed.  The parameter estimates (and 95% confidence intervals) from the model fits are shown in ",tblcaps("BonitoRes",display='cite'),"."))
```

tblcaps("BonitoRes",paste0("Parameter estimates, residual sums-of-squares (RSS), and Akaike Information Criterion (AIC) from the fits of ",eqncaps('SomersMod',display='cite')," and ",eqncaps('PaulyMod2',display='cite')," to the Bonito data."))
```

```{r echo=FALSE, comment=""}
cat("Somers Model (Equation 3) fit")
print(round(cfBon1,2),na.print="-")
cat("Modified Pauly Model (Equation 6) fit")
print(round(cfBon2,2),na.print="-")
```

@@

>>>>

---
title: "Revisiting the von Bertalanffy Seasonal Cessational Growth Function of Pauly et al. (1992)"
author: "Derek H. Ogle"
date: '`r format(Sys.time(), "%A, %b %d, %Y")`'
output:
  pdf_document:
    fig_height: 3.5
    fig_width: 4.5
  html_document:
    fig_height: 3.5
    fig_width: 4.5
csl: american-fisheries-society.csl
bibliography: SeasonalGrowth.bib
---

@@

# setup figure, table, and equation captioning
library(captioner)
figcaps <- captioner(prefix="Figure")
tblcaps <- captioner(prefix="Table")
eqncaps <- captioner(prefix="Equation")
eqncaps("vbMain")
eqncaps("vbMain2")
eqncaps("SomersMod")
eqncaps("PaulyMod")

@@

----

# Version notes

1. Likely will not include simulated data results at the bottom.  They are a weak test at best.
1. Would like to add 1-3 more examples of real data with varying seasonal effects (clear cessation (like Bonito), clear cessation with decline in mean length-at-age, and no clear cessation).
1. The citation styles are not correct in this automated version.

----

@@

The mean length-at-age for many fish [@Haddon2011] and other aquatic animals (e.g., @Hota1994; @Harwoodetal2014) is often modeled with the von Bertalanffy growth function (VBGF; @vonBertalanffy1938).  The parameterization of the VBGF attributable to @BevertonHolt1957 is most common and may be expressed as

@@

Many fish exhibit seasonal oscillations in growth as a response to seasonal changes in environmental factors such as temperature, light, and food supply (e.g., @Bayley1988; @Paulyetal1992; @Baconetal2005; @GarciaBerthouetal2012; @CarmonaCatotetal2014).  `r eqncaps("vbMain2",display="cite")` of the traditional VBGF has been modified, usually with a sine function, to model these seasonal oscillations in growth.  The most popular of these modifications, from @HoenigChoudarayHanumara1982 and @Somers1988, is

@@

where $C$ modulates the amplitude of the growth oscillations and corresponds to the proportional decrease in growth at the depth of the oscillation (i.e., "winter"), and $t_{s}$ is the time between time 0 and the start of the convex portion of the first sinusoidal growth oscillation (i.e., the inflection point).  If $C$=0, then there is no seasonal oscillation and `r eqncaps("SomersMod",display="cite")` reduces to  `r eqncaps("vbMain2",display="cite")` and the typical VBGF (`r figcaps("SomersExPlot",display="cite")`).  If $C$=1, then growth completely stops once a year at the "winter-point" ($WP$), whereas values of 0&lt;$C$&lt;1 result in reduced, but not stopped, growth during the winter (`r figcaps("SomersExPlot",display="cite")`).  Note that $WP=t_{s}+\frac{1}{2}$ because the sine function in `r eqncaps("SomersMod",display="cite")` has a period (i.e., the growth period) of one year.  Some confusion has surrounded the use of `r eqncaps("SomersMod",display="cite")`, although @GarciaBerthouetal2012 carefully clarified its form.

@@

Values of $C$&gt;1 (or &lt;0) in  `r eqncaps("SomersMod",display="cite")` allow seasonal decreases in mean length-at-age (`r figcaps("SomersExPlot",display="cite")`).  A decrease in mean length is unlikely for organisms whose skeletons largely preclude shrinkage [@Paulyetal1992], although a seasonal decrease in mean length-at-age is possible if size-dependent overwinter mortality occurs [@GarciaBerthouetal2012].  @Paulyetal1992 modified `r eqncaps("SomersMod",display="cite")` to include a true seasonal no-growth period where mean length was not allowed to decrease and smoothly transitioned into and out of the no-growth period.  Specifically, their modification is

@@

@Paulyetal1992 devised  `r eqncaps("PaulyMod",display="cite")` by assuming $C$=1 and replacing $2\pi$ in `r eqncaps("SomersMod",display="cite")` with $\frac{2\pi}{1-NGT}$ (i.e., restricting the seasonal oscillation to the growth period and noting that $K^{'}$ only operates during the growth period).  Their modification may be described geometrically (though not algorithmically) in two steps.  First, `r eqncaps("SomersMod",display="cite")` with (fixed) $C$=1 is fit to the observed lengths and ages that have had the cumulative $NGT$ subtracted (i.e., using $t^{'}$).  This growth trajectory is then separated at each $WP$ and horizontal segments that are $NGT$ units long are inserted at these points.  This forms a growth trajectory that smoothly transitions into and out of the no-growth periods (`r figcaps("PaulyExPlot",display="cite")`).

@@

@Paulyetal1992 provided a "diskette" that contained a computer program to estimate the parameters of `r eqncaps("PaulyMod",display="cite")`.  The diskette is difficult (at best) to obtain and the source code is no longer available (D. Pauly, pers. comm.).  @Paulyetal1992 did describe the operations performed by their program, but there is no description of how $t^{'}$ was operationalized.  This is an important step in using `r eqncaps("PaulyMod",display="cite")` because $t^{'}$ is a function of $t$, but it is also a function of $NGT$ and $t_{s}$, which are parameters to be estimated during the model-fitting process.  Thus, the values for $t^{'}$ change with each iteration of the non-linear model-fitting algorithm.

@@

As noted by @Paulyetal1992 the calculation of $t^{'}$ depends on the observed age ($t$) and the cumulative no-growth time prior to $t$.  In practice, the calculation of $t^{'}$ also depends on the position of the no-growth period within a year.  Here, the position of the no-growth period is defined relative to $WP$ and $NGT$, such that the following algorithm may be used to convert from observed ages ($t$) to ages adjusted for cumulative $NGT$ prior to age $t$ ($t^{'}$).

1. Shift the age ($t$) by subtracting the start of the no-growth ($SNG$) period (i.e., $SNG=WP-\frac{NGT}{2}$$=t_{s}+\frac{1}{2}-\frac{NGT}{2}$) from $t$, such that a whole number will represent the start of a no-growth period.  For example, if $SNG$=0.4, then $t$=2.4 will become 2.0 and $t$=2.9 will become 2.5.
1. Subtract the whole number age from the shifted age from Step 1 such that the remaining decimal represents the fraction of a shifted year.  For example, a 0 will result if the shifted age is 2.0 and a 0.5 will result if the shifted age is 2.5.
1. Substract the $NGT$ from the value from the previous step.
1. If the value from the previous step is negative, then the age is within the no-growth period and the negative value should be replaced with a zero.  Otherwise, the positive value represents the amount of time into a growth period.
1. Add the value from the previous step to the total growth time completed (i.e., the product of the number of growth periods completed and the length of the growth period ($1-NGT$)).
1. Compute $t{'}$ by adding back the $SNG$ that was subtracted in Step 1.

@@

# Fitting the Function
## Real Data

@Stewartetal2013 examined the growth of 215 Bonito (*Sarda australis*) sampled from commercial landings.  Fork lengths (mm) were measured for each fish and decimal ages were recorded as the number of opaque zones observed on otolith thin sections plus the proportion of the year after the designated birthdate (see @Stewartetal2013 for more detailed methods).  @Stewartetal2013 fit `r eqncaps("SomersMod",display="cite")` to these data but constrained $C$ to not exceed 1.  Their model fit resulted in the boundary condition of $C=1$, which suggested that Bonito ceased to grow at *at least* one point.  This result suggests that `r eqncaps("PaulyMod",display="cite")` may fit these data better and allow the length of the no-growth period to be estimated [@Paulyetal1992].

`r eqncaps("PaulyMod",display="cite")` fit the Bonito data slightly better than `r eqncaps("SomersMod",display="cite")` with slightly lower residual sums-of-squares (RSS) and Akaike Information Criterion (AIC) values.  The length of the no-growth period was estimated to be `r formatC(cfBon2["NGT","Est"],format="f",digits=3)` or `r formatC(cfBon2["NGT","Est"]*100,format="f",digits=1)`% of the year.  The $t_{s}$ parameters were equal and the $L_\infty$ parameters were similar, but the $t_{0}$ parameters differed somewhat between the two models (`r tblcaps("BonitoRes",display="cite")`).  The $K$ from `r eqncaps("SomersMod",display="cite")` was equal to $K'$ from `r eqncaps("PaulyMod",display="cite")` multiplied by $1-NGT$ (`r tblcaps("BonitoRes",display="cite")`).  Graphically, there was little perceptual difference in the model fits (`r figcaps("BonitoRes",display="cite")`).

```{r echo=FALSE, results="hide"}
## A plot
par(xaxs="i",yaxs="i",mar=c(3,3,0.6,0.6),mgp=c(1.7,.5,0),tcl=-0.3,las=1)
plot(FL~Age,data=bon,pch=19,col=col2rgbt("black",1/4),
     xlab="Age (years)",ylab="Fork Length (mm)",xlim=c(0,4),ylim=c(0,70))
curve(VBSCGF(x,coef(fitBon2)),from=0,to=4,lwd=4,add=TRUE)
curve(vbSO(x,coef(fitBon1)),from=0,to=4,lwd=2,add=TRUE,col="gray50")

figcaps("BonitoRes",paste0("Fork length at age for Australian Bonito with the best-fit of ",eqncaps('PaulyMod',display='cite')," (black line) and ",eqncaps('SomersMod',display='cite'),", with fixed C=1, superimposed (gray line).  The parameter estimates (and 95% confidence intervals) from the model fits are shown in ",tblcaps("BonitoRes",display='cite'),"."))
```

tblcaps("BonitoRes",paste0("Parameter estimates, residual sums-of-squares (RSS), and Akaike Information Criterion (AIC) from the fits of ",eqncaps('SomersMod',display='cite')," and ",eqncaps('PaulyMod',display='cite')," to the Bonito data."))
```

```{r echo=FALSE, comment=""}
cat("Somers Model (Equation 3) fit")
print(round(cfBon1,2),na.print="-")
cat("Modified Pauly Model (Equation 4) fit")
print(round(cfBon2,2),na.print="-")
```

## Simulated Data
Simple simulated data are used initially to demonstrate the fitting of `r eqncaps("PaulyMod",display="cite")`.  The data were simulated by randomly selecting 200 real numbers from a uniform distribution between 0 and 5 to serve as observed ages, plugging these ages into `r eqncaps("PaulyMod",display="cite")` to compute a mean length for each age, and then adding a random deviate from a normal distribution with a mean of 0 and a standard deviation of $\sigma$ to each mean length to simulate observed individual lengths.  Eight data sets were simulated for all combinations of "early" ($t_{s}=0.2$) and "late" ($t_{s}=0.8$) maximum growth points, "short" ($NGT=0.2$) and "long" ($NGT=0.5$) no-growth periods, and "low" ($\sigma=0.3$) and "high" ($\sigma=0.5$) individual variabilities.

The `nls()` function from R was then used to estimate parameter values for the nonlinear `r eqncaps("PaulyMod",display="cite")` fit to each simulated data set.  The "port" algorithm was used so that $L_{\infty}$ and $K^{'}$ could be constrained to be positive and $t_{s}$ and $NGT$ could be constrained to be between 0 and 1.  The estimated parameters for each data were consistently very close to the values used to create the data set, with the possible exception of $t_{0}$.  One example, for a "long" $NGT$, "late" $t_{s}$, and low individual variability, is shown in `r figcaps("PaulySimRes",display="cite")`.  The results from fitting `r eqncaps("PaulyMod",display="cite")` to these data sets are not surprising as one would expect the parameter estimates to be very close to the values used to create them, given that `r eqncaps("PaulyMod",display="cite")` was used to both create and fit the data.

@@

tblcaps("PaulySimRes",paste0("Parameter estimates from fitting ",eqncaps('PaulyMod',display='cite')," to simulated data sets using $L_{\\infty}$=",Linf,", $K'$=",Kpr,", $t_{0}$=",t0,", and varying values of $t_{s}$, $NGT$, and $\\sigma$."))
```

@@

@@

# Acknowledgments

John Stewart (New South Wales Department of Primary Industries Fisheries) graciously provided the Bonito length-at-age data.
